---
title: "SISMID_POMP_SIM"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### loading packages and dataset
```{r}

library(tidyverse)
library(doParallel)
library(doRNG)
registerDoParallel()
registerDoRNG(2488820)
read_csv(paste0("http://kingaa.github.io/sbied/stochsim/", "Measles_Consett_1948.csv")) %>%
  select(week, reports=cases) -> meas # change colname cases to reports

meas %>%  as.data.frame() %>% head() # view the meas data as dataframe
```

## POMP model in R script
**components:**
  - rprocess
  - rmeasure
  - optional: dmeasure
  
**For DEMONSTRATION ONLY**
- slower in computational time
- code written in Csnippt speed up the process


### set up one step transition simulator for the unoberserved state process
    - simple counting process, f(Xt|Xt-1), rprocess
      - count one event at a time (t), in this case, a **stochastic process**
    - **Euler's method**:approximation (numerical) solution of ODE function 
      - stepping foward in time increments, Delta.t, where Delta.t < t
    - S(t), I(t), R(t), **unobserved state variables**
    - rho, probability of reporting 
    - H, number of reported cases (for measles, much lower transmission rate after reporting, thus relate to I->R)
      - **accumlator variable**, time unit in t, instead of delta.t
        - reset to 0 at beginning of each week
```{r}
# # set up in R
# # set up transition step into a function
# sir_step <- function(S,I,R,H,N,Beta, mu_IR, delta.t,...){
# 
#   # Euler method chosen, binomial approximation with exponential transition  probability, Binomial[X=aprox(S(t)), p=1-exp{-Mu_SI(approx(I(t))delta.t)}],where p is the probability of one unit of susceptible moving to the infected class
# 
#   # rate of changes between compartments, counting process, which is the flow between compartments
#   dN_SI <- rbinom(n=1, size=S, prob=1-exp(-Beta*I/N*delta.t)) # Delta_N_SI, S->I, over Delta.t
#   dN_IR <- rbinom(n=1, size = I, prob=1-exp()) # Delta_N_IR, I->R, over Delta.t
# 
#   # Compartments
#   S <- S - dN_SI # S(t)
#   I <- I + dN_SI -dN_IR # I(t)
#   R <- R + dN_IR # R(t)
#   H <- H + dN_IR
#   c(S = S, I = I, R = R) # combine the returning results into vector c()
#   }

```

### Initial status, day 0
```{r}
# # eta, fraction of N that are susceptible
# sir_rinit <- function(N, eta, ...){
# 
#   c(S= round(n * eta), I = 1, R = round(N*(1-eta)))
# 
# }

```

---


### Define rmeasure and dmeasure
    - since our data, reports, is the results of reported cases
    - model data, f(Yt|Xt),into a binomial process
      - reports_t ~ Binomial[X= H(t), p= rho], where rho is the probability of reporting
```{r}
# 
# # dmeasure, Evaulating the density of f(Yt|Xt) at report given Xt = H(t)
# # report log function, if specify= TRUE
# sir_dmeas <- function(reports, H, rho, log, ...){
#   dbinom(x=reports, size=H,prob=rho, log=log) # report the evaulation results in log
# }
# 
# # rmeasure, drawing from the density distribution, f(Yt|Xt) given Xt=H(t)
# sir_rmeas <- function(H, reports, ...){
#   c(reports=rbinom(n=1, size=H, prob=rho))
# }


```

## POMP model

### Building POMP model
- using **accumvars** arguments to specify **accumulator variable**
```{r}
# 
# meas %>% 
#   pomp(
#     times = "week", t0=0, # t = week
#     rprocess = euler(sir_step,delta.t = 1/7), # delta.t = day
#     rinit=sir_rinit,
#     accumvars = "H",
#     rmeasure=sir_rmeas,
#     dmeasure=sir_dmeas
#     ) -> measSIR

```

---

## POMP model in Csnippt
   - speed up the computational time
   
### Write all components in Csnippts
```{r}
sir_step <- Csnippet("
                     double dN_SI = rbinom(S, 1-exp(-Beta*I/N*dt));
                     double dN_IR = rbinom(I, 1-exp(-mu_IR * dt));
                     S -= dN_SI;
                     I +=dN_SI - dN_IR;
                     R += dN_IR;
                     H += dN_IR; 
                     ")

sir_rinit <- Csnippet("
                      S = nearbyint(eta * N);
                      I = 1;
                      R = nearbyint((1-eta) * N);
                      ")

sir_dmeas <- Csnippet("
                      lik = dbinom(reports, H, rho, give_log);
                      ")

sir_rmeas <- Csnippet("
                      reports = rbinom(H, rho);
                      ")


```


### Define POMP model with Csnippts

```{r}
library(pomp)

meas %>% 
  pomp(
    times = "week", t0=0, # t = week
    rprocess = euler(sir_step,delta.t = 1/7), # delta.t = day
    rinit=sir_rinit,
    accumvars = "H",
    rmeasure=sir_rmeas,
    dmeasure=sir_dmeas,
    statenames = c("S", "I","R","H"), # unobserved state variable stated in Csnippet
    paramnames = c("Beta", "mu_IR","N","eta","rho") # parameter values need to specify
  ) -> measSIR

measSIR


```

### Simulate
- start with choosing plausible parameter values

```{r}
measSIR %>% 
  simulate(params=c(Beta=7.5,mu_IR=0.5,rho=0.5,eta=0.03,N=38000),
           nsim=20,format="data.frame",include.data=TRUE
           )-> sims

# blue: data, red = simulation
sims %>% 
  ggplot(aes(x=week,y=reports,group=.id,color=.id == "data"))+
  geom_line()+
  guides(color=FALSE)
```

## Exercise: SEIR model
- add a E compartment
```{r}
seir_step <- Csnippet("
  double dN_SE = rbinom(S,1-exp(-Beta*I/N*dt));
  double dN_EI = rbinom(E, 1-exp(-mu_EI*dt));
  double dN_IR = rbinom(I,1-exp(-mu_IR*dt));
  S -= dN_SE;
  E += dN_SE - dN_EI;
  I += dN_EI - dN_IR;
  R += dN_IR;
  H += dN_IR;
  ")

seir_rinit <- Csnippet("
  S = nearbyint(eta*N);
  E = 0;
  I = 1;
  R = nearbyint((1-eta)*N);
  H = 0;
  ")

seir_dmeas <- Csnippet("
  lik = dbinom(reports,H,rho,give_log);
  ")

seir_rmeas <- Csnippet("
  reports = rbinom(H,rho);
  ")

# excercise 2: changed from a SIR model to a SEIR model 
#   added parameter "mu_EI" into "paramnames" vector
# added "E" into "statenames" vector
meas %>%
  pomp(
      times = "week", t0=0, # t = week,
       rprocess=euler(seir_step,delta.t=1/7),
       rinit=seir_rinit,
       rmeasure=seir_rmeas,
       dmeasure=seir_dmeas,
       accumvars="H",
       statenames=c("S","I","E","R","H"),
       paramnames=c("Beta","mu_IR","mu_EI","N","eta","rho")
       ) -> measSEIR

measSEIR %>% 
  simulate(params=c(Beta=7.5,mu_IR=0.5,mu_EI=0.66,rho=0.5,eta=0.03,N=38000),
           nsim=20,format="data.frame",include.data=TRUE
           )-> sims_seir

# blue: data, red = simulation
sims_seir %>% 
  ggplot(aes(x=week,y=reports,group=.id,color=.id == "data"))+
  geom_line()+
  guides(color=FALSE)

```